White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-07-11 papers/wdmerger_II/paper.tex
Add some data to paper II

\author{Max P. Katz\altaffilmark{1}}
\author{Michael Zingale\altaffilmark{1}}
\author{Alan C. Calder\altaffilmark{1,2}}
\author{F. Douglas Swesty\altaffilmark{1}}
\author{Ann S. Almgren\altaffilmark{3}}
\author{Weiqun Zhang\altaffilmark{3}}

@@

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}.
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. We choose by default $f_{X} = f_{e} = 0.1$, 
a choice which is slightly smaller than others in the literature. (Note that previous 
collision papers we are aware of only limited based on changes in internal energy, 
not on cahnges in species.)  The sensitivity of results to this choice will be discussed in 
\autoref{sec:2D:timestep}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

@@

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. We
too have developed an option for our burner to do this,
the ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied. To achieve this we directly multiply the
right-hand-side vector in the integration by a constant factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $t_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $t_e$ to be the value of the
internal energy of the zone at the beginning of the burn. If
instead one allowed $c_s$ and $e$ to evolve with the burn, one
would obtain a less conservative limiter in the case of explosive
burning, as $c_s$ and $e$ are both increasing in this case.
As the point of the limiter is to ensure that the changes to the
\textit{original} energy are small enough so that the following
hydrodynamics update can advect away newly generated energy
quickly enough to avoid a nuerically seeded detonation,
we desire the most conservative version of the limiter. We discuss
our results with the suppressed burning mode in \autoref{sec:2D}.

@@

For all the 3D simulations to follow, we use the same coarse grid as in Paper I:
there are 256 zones per dimension, and the width of each axis is $1.024 \times 10^{10}$
cm, yielding a base resolution of 400 km. When we use adaptive-mesh refinement, we will
use the refinement scheme described in \autoref{sec:unstable_burning}. This is
acceptable because the outcome of the collision problem is chiefly a function
of the propagation of burning front through the WDs, so for this problem one
can take some liberties in other parts of the algorithm for the purpose of
computational efficiency, which can always be turned back on later if a more
accurate answer is desired. So, for example, we do not need to add refinement
everywhere in the WDs in the interest of getting the collision dynamics to be
slightly more accurate. Additionally, we choose to solve the Poisson equation
for the self-gravity of the system only on the coarse grid. The gravitational
potential and acceleration on the finer levels are interpolated from the coarse
grid.

@@

%==========================================================================
% 2D simulations
%==========================================================================
\section{Two-Dimensional Simulations}
\label{sec:2D}

While white dwarf collisions are three-dimensional events, there is much to
be learned from simpler, and much computationally cheaper, two-dimensional
axisymmetric collision simulations. These are implemented in the cylindrical
coordinate system described in \autoref{sec:problemsetup}.
We simulate collisions in 2D cylindrical coordinates for three purposes.
First, we can quickly get up to speed on the nature of the problem and then
benchmark existing and new code tools such as our nuclear postprocessing analysis.
Second, in the regime where the 2D and 3D simulations can be usefully compared, 
zero-impact parameter head-on simulations ($b = 0$), we can understand to
what extent the reduced-dimensionality approach can reproduce the 3D results
and thus determine whether 2D simulations can be used for obtaining scientific
predictions, including nucleosynthesis yields and observable quantities.
Third, we can cheaply test the effect of various code choices such as the number
of isotopes in the nuclear network; we can afford to run the same problem many
times while tweaking parameters one by one to learn their effects. 
Prepared with these results and a better understanding of how reliable our
approximations are, we can then proceed to performing 3D simulations with
our code choices already made and with confidence that we understand the
uncertainty resulting from these choices.

\subsection{Dynamics}
\label{sec:2D:dynamics}

First we briefly consider whether the dynamical evolution of the system
is consistent with simple physics we understand. We ask two questions.
First, is the time it takes for the white dwarfs to collide consistent with
basic kinematics? Second, is the amount of energy released in the
collision by the nuclear reactions comparable to the amount of kinetic
energy we infer from observations of Type Ia supernovae? We must ensure
that we are getting these basics right before we can advance to more
complicated questions. This is not necessarily a trivial matter. We have
observed, for example, that at low resolution, the WDs take longer to
collide than at higher resolution, because the gravitational field is
much less accurate.

\subsection{Timestep Limiting}
\label{sec:2D:timestep}

Here we discuss the effect of limiting the hydrodynamic timestep based on
changes in the internal energy in the nuclear reaction network.

\subsection{Nuclear Network}
\label{sec:2D:network}

Now we move to a discussion of the effect of including various isotopes
in the nuclear reaction network. In particular we compare the \aproxthirteen,
\aproxnineteen, and \aproxtwentyone reaction networks.

\subsection{Nucleosynthetic Postprocessing}
\label{sec:2D:postprocessing}

We consider also the results of our postprocessing step for nucleosynthesis
yields, to understand whether the results for the isotopes not included in
our networks make sense.

%==========================================================================
% 3D simulations
%==========================================================================
\section{Three-Dimensional Simulations}
\label{sec:3D}

Now that we have analyzed the various parts of our simulation software and
discussed our code choices, we consider full 3D simulations to understand
the potential for white dwarf collisions to be progenitors of Type Ia
supernovae. We focus on two main quantities of interest: the total amount of
nickel produced in the explosion, which directly powers the light curve
observed in SNe Ia; and, the gravitational wave signature produced by
these events.

>>>>

\author{Max P. Katz\altaffilmark{1}}
\author{Michael Zingale\altaffilmark{1}}
\author{Alan C. Calder\altaffilmark{1,2}}
\author{F. Douglas Swesty\altaffilmark{1}}
\author{Ann S. Almgren\altaffilmark{3}}
\author{Weiqun Zhang\altaffilmark{3}}
\author{Frank X. Timmes\altaffilmark{4,5}}

@@

\altaffiltext{4}
{
  The Joint Institute for Nuclear Astrophysics, USA
}

\altaffiltext{5}
{
  School of Earth and Space Exploration, Arizona State University, Tempe, AZ, USA
}

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. We choose by default $f_{X} = f_{e} = 0.1$, 
a choice which is slightly smaller than others in the literature. (Note that previous 
collision papers we are aware of only limited based on changes in internal energy, 
not on cahnges in species.)  The sensitivity of results to this choice will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

@@

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. We
too have developed an option for our burner to do this,
the ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied. To achieve this we directly multiply the
right-hand-side vector in the integration by a constant factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $t_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $t_e$ to be the value of the
internal energy of the zone at the beginning of the burn. If
instead one allowed $c_s$ and $e$ to evolve with the burn, one
would obtain a less conservative limiter in the case of explosive
burning, as $c_s$ and $e$ are both increasing in this case.
As the point of the limiter is to ensure that the changes to the
\textit{original} energy are small enough so that the following
hydrodynamics update can advect away newly generated energy
quickly enough to avoid a nuerically seeded detonation,
we desire the most conservative version of the limiter. We discuss
our results with the suppressed burning mode in \autoref{sec:parameters}.

@@

For all the 3D simulations to follow, we use the same coarse grid as in Paper I:
there are 256 zones per dimension, and the width of each axis is $1.024 \times 10^{10}$
cm, yielding a base resolution of 400 km. When we use adaptive-mesh refinement, we will
use the refinement scheme described in \autoref{sec:unstable_burning}. This is
acceptable because the outcome of the collision problem is chiefly a function
of the propagation of burning front through the WDs, so for this problem one
can take some liberties in other parts of the algorithm for the purpose of
computational efficiency, which can always be turned back on later if a more
accurate answer is desired. So, for example, we do not need to add refinement
everywhere in the WDs in the interest of getting the collision dynamics to be
slightly more accurate. Additionally, we choose to solve the Poisson equation
for the self-gravity of the system only on the coarse grid. The gravitational
potential and acceleration on the finer levels are interpolated from the coarse
grid. This grants enhanced computational speed without a serious effect on the
accuracy of the simulation (for the case where the fine grids cover the stars,
we have checked that the more accurate gravitational forcing would only modify
slightly the time to impact and the subsequent detonation process).

@@

As in Paper I, the region outside the WDs is filled with a low-density ambient
material with $\rho = 10^{-4}\ \text{g / cm}^{-3}$ whose composition is the
same as that of the WDs (which is equal carbon and oxygen by mass, unless
stated otherwise). Reactions in this region are unimportant, so for
computational efficiency, we disable all nuclear burning for zones that have
$\rho < 10^6\ \text{g / cm}^{-3}$ and $T < 10^8\ \text{K}$. The initial ambient
temperature is $10^7\ \text{K}$, which is the same as the initial temperature
everywhere throughout the WDs, and we set the temperature floor to the same
value (any zone that reaches below this temperature will be reset).

\castro\ uses the dual energy formalism of \cite{bryan:1995,ENZO}, where we evolve equations
for both the total energy $E$ and the internal energy $e$, using one or the other to
determine the pressure in the hydrodynamics step (if we are using the total energy, we
compute the corresponding internal energy by subtracting the kinetic energy $K$). The switching
parameter is $\eta_1 = 10^{-3}$, and if $(E - K) / E < \eta_1$, we use the internal energy variable $e$
in the equation of state call that obtains the pressure. Otherwise we use $(E - K)$. The
idea behind this is that in high Mach number flows where the kinetic energy dominates the
total energy, we do not want to risk generating a negative estimate of the internal energy
due to roundoff error. Similarly, at regular intervals when we enforce consistency of the
internal energy variable (which occurs a few times per timestep), we sync the internal energy
variable so that it is equal to $E - K$, skipping this step only if $(E - K) / E < \eta_2$.
We adopt by default $\eta_2 = 10^{-4}$. This choice can matter much more for reacting
hydrodynamics than non-reacting hydrodynamics. For example, a collision can only lead to a
detonation if there is sufficient conversion of kinetic to thermal energy. Obviously
this happens physically, but if the parameter $\eta_2$ is set high enough then it could
in principle interfere by not allowing the simulation to efficiently convert kinetic energy
to thermal energy. (This does not end up mattering much for our simulations because the
kinetic energy does not strongly dominate the internal energy for colliding WDs.) And on
a similar note, when using this formalism it is ambiguous which variable to use ($(E - K)$
or $e$) when calculating the initial temperature that goes into the reaction step. We
have introduced a new parameter $\eta_3$ that uses the same logic as the first two: if
$(E - K) / E < \eta_3$, we use $(E - K)$ in the equation of state call that obtains the
temperature; otherwise, we use $e$. However, unlike the other two, the \castro\ default
is $\eta_3 = 1.0$, meaning that we normally use the internal energy variable.

%==========================================================================
% Parameter Study
%==========================================================================
\section{Parameter Study}
\label{sec:parameters}

Before turning to the effect that adaptive mesh refinement has on WD collisions,
we will first examine the effect of a number of other code parameters. For most
of the following we will use two-dimensional axisymmetric simulations which have
impact parameter $b = 0$, unless otherwise specified, and we will stick to a
moderate resolution (we use only the coarse grid described in \autoref{sec:problemsetup}).
For all simulations in this section, we use a pair of $0.64\ \msolar$ WDs, a binary
system that has been studied in most of the collision papers to date. This allows
us to cheaply test how the collision responds to many effects. One limitation of
this is that we will not find out exactly how a simulation differs in the case of a
high-resolution 3D run when these parameters changed. Another is that we will not
perform a full parameter study where we consider the full non-linear dependence of
all parameters on all other parameters; here we are merely interested in qualitative
trends that will help to understand the limitations of WD collision simulations.

With those caveats in mind, let us consider the tests and the results we obtained.
Our primary metric for the following tests will be the amount of $^{56}$Ni generated
in the collision, as this is the parameter most directly related to the observable
quantities of interest for Type Ia supernovae. Specifically, we collect information
at the end of every timestep about the total amount of nickel on the grid (in solar
masses), and we will use the maximum value of this nickel mass. When the stopping
criterion described in \autoref{sec:problemsetup} is used, this nickel mass is generally
constant or decreasing by the end of the run. The typical amount of nickel mass generated
for our runs at low resolution is $0.2\ \msolar$, which is significantly lower than
that obtained by \cite{raskin:2010} and \cite{kushnir:2013} for the same problem.
We will discuss this in \autoref{sec:resolution}.

\subsection{Timestepping}
\label{sec:parameters:timestepping}

For the standard self-heating burns that we use, the most important parameters we
examined come from the timestep limiting scheme described in \autoref{sec:hydrocoupling}.
The parameters $f_{be}$ (from \autoref{eq:timestep_e}) and $f_{bX}$ (from
\autoref{eq:timestep_X}) control the size of the timestep as a function of the
burning rate. As the timestep gets smaller, the error due to the Strang splitting
scheme coupling the reactions and the burning also gets smaller, and we expect
that the results become more accurate.
\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.425]{{{plots/dtnuc_max_Ni56_m_P_0.64_m_S_0.64}}}
  \caption{Nickel production as a function of the timestep factors $f_{be}$
           (solid blue, limiting the timestep on changes in internal energy) and
           $f_{bX}$ (dashed green, limiting the timestep on changes in species).
           For a given $f_{be}$, $f_{bX}$ with the same value typically limits
           the timestep by at least an order of magnitude more.
           \label{fig:timestep_factor}}
\end{figure}
\autoref{fig:timestep_factor} demonstrates the effect of both parameters on the nickel
production. As either $f$ becomes small, the restriction on the timestep becomes quite
severe: for $f_{bX} = 0.01$, the minimum timestep in the simulation is smaller than
$2 \times 10^{-9}\ \text{s}$ and the full simulation requires over 750,000 timesteps.
Yet it is clear that even for this steep value of the limiter, which is much stricter
than the limiter used in, e.g., \cite{hawley:2012} and \cite{raskin:2010}, the
simulation has not fully converged in nickel production. These limiters are somewhat
redundant in the sense that a given $f_{bX}$ corresponds to another (higher) $f_{be}$,
as is evident by the fact that on the graph the curves look similar in shape but with
a horizontal offset.

We also examined the effect of the method used to enforce the limiting, using the
four limiting modes discussed in \autoref{sec:hydrocoupling}. The results are shown
in \autoref{table:burninglimitermode} for the case $f_{be} = 0.3$ (we did not use
limiting with $f_{bX}$ for this test). The two modes that use the full last timestep's
worth of information to estimate $\dot{e}$ yield a slightly larger nickel mass than the
two modes that only use more recent information, but the difference is negligible.

\input{plots/burning_limiter_mode_m_P_0.64_m_S_0.64.tbl}

%TODO: add a ``Nonaka plot''

For all remaining tests in this section, unless otherwise stated, we take $f_{be} = 10^2$
and disable limiting with $f_{bX}$ by setting it to a large number. This choice is a balance
between accuracy and efficiency: we avoid the obviously wrong answers that occur for $f_{be} > 10^3$
while ensuring that the timesteps are long enough that the following tests are computationally
inexpensive.

\subsection{Nuclear Network}
\label{sec:parameters:network}

Now we move to a discussion of the effect of including various isotopes
in the nuclear reaction network. In particular we compare the \isoseven, \aproxthirteen,
\aproxnineteen, and \aproxtwentyone\ reaction networks. In all cases we run
the same problem setup, and the unused isotopes (everything but carbon and oxygen)
have their mass fractions in every zone set to a negligibly small number at initialization.

\subsection{Burning Mode}
\label{sec:parameters:burningmode}

\subsection{Other Parameters}

We tested a few other parameters that ended up having no serious effect on the outcome.
The dual energy parameter $\eta_2$ described in \autoref{sec:problemsetup} can have a
minor effect on the nickel production, yielding variations on the order of a few percent,
but there is no clear pattern. The dual energy parameter $\eta_3$ has no effect on the
outcome: the answer is the same for $\eta_3 = 0$ and $\eta_3 = 1$, which makes sense
because all of the reacting regions have a low enough kinetic energy relative to their
total energy such that the internal energy variable is always kept in sync with $(E - K)$.
(We did not look at $\eta_1$ because that has a direct effect only on the hydrodynamics,
not the reactions.) We also tested whether our choice for the temperature floor mattered
by lowering it to $10^6\ \text{K}$ and $10^5\ \text{K}$, and this had no effect on nickel
production. It does allow parts of the WDs to become colder as they move, due to numerical
errors in the advecting flow (the Helmholtz equation of state is very insensitive to
temperature, so even small changes in the internal energy can create very large changes
in the implied temperature), but the amount of energy injected by the collision is so
large that the final result is insensitive to whatever the initial temperature of the
WDs were. We also checked whether there was any consequence of our choice to disable the
burning for small enough $\rho$ and $T$. Lowering the minimum $\rho$ for reactions to
$10^0\ \text{g / cm}^{-3}$ from $10^6\ \text{g / cm}^{-3}$ had a negligible effect,
and varying the minimum $T$ for reactions from $2 \times 10^{7}\ \text{K}$ to
$4 \times 10^{8}\ \text{K}$ also had no meaningful effect.

%TODO: maybe show an image of the temperature field in the WDs before colliding to demonstrate the point about numerical error?

The choice of the WD composition to be equal carbon and oxygen by mass is a common
choice in the literature and is broadly comparable to what stellar evolution codes
yield for WDs of this mass. Still, it is an arbitrary approximation, especially the
insistence that the composition is uniform throughout the WD. In \autoref{table:co}
we list the nickel production as a function of the initial carbon mass fraction.
This can have a few percent impact on the outcome for the plausible values we tested.
In future work we may look at the effect of a more realistic progenitor structure,
including the effects of a helium shell on the surface of the WDs \citep{holcomb:2015}
and a non-uniform interior structure.

\input{plots/co_frac_m_P_0.64_m_S_0.64.tbl}

%==========================================================================
% Resolution Dependence
%==========================================================================
\section{Resolution Dependence}
\label{sec:resolution}

%==========================================================================
% Gravitational Wave Signature
%==========================================================================
\section{Gravitational Wave Signature}
\label{sec:gwsignature}
